\documentclass[11pt]{article}
\usepackage[a4paper,margin=1in]{geometry}
\usepackage{microtype}
\usepackage{lmodern}
\usepackage{amsmath,amssymb}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{siunitx}
\usepackage{hyperref}
\usepackage[round,authoryear]{natbib}
\usepackage{xurl}
\usepackage{caption}

\hypersetup{colorlinks=true,linkcolor=black,citecolor=black,urlcolor=blue}
\bibliographystyle{plainnat}

\title{Pre-Registered Cross-Regional Earthquake Precursor Evaluation:\\
A Five-Failure-Mode Catalogue and Cross-Regional Confirmation of\\
Pervasive Foreshock Activity at AUC$\,=\,$0.70 / $z\,=\,8$\\
with Quantitative $M$-Scaling}
\author{Matthew Loftus\\
\small Cedar Loop LLC, Bremerton, WA, USA\\
\small \texttt{matthew.a.loftus@gmail.com}}
\date{\today}

\begin{document}
\maketitle

\begin{abstract}
We pre-registered a CSEP-style cross-regional earthquake-precursor
evaluation across six training and two test regions, six features, and
three null models, and ran it under the locked protocol with one
documented amendment. The protocol surfaced \textbf{five concrete
methodological failure modes} that the field has not previously
catalogued in this combination: (i)~the inherited overlap-rejection rule
collides with itself in event-dense subduction zones (Japan, Chile, Mexico,
Alaska all drop); (ii)~ANSS ComCat is teleseismic for non-US regions,
forcing $M_c$ to the target threshold; (iii)~the Aki MLE $b$-value
$N{\ge}30$ floor combined with 30-day windows at moderate $M_c$ renders
the $b$-feature uncomputable in three of four qualifying regions;
(iv)~both pre-registered test regions fail (i)~and are structurally
inaccessible; (v)~the pre-registered precursor window $[t-30,t)$ includes
the foreshock period, conflating long-distance precursor with
foreshock-detection. Per-window scalar AUCs and a joint logistic-regression
$+$ random-forest classifier across the six features all yield clean
nulls. A TLS-style template-correlation scan over the Benioff strain
trajectory passes the pre-registered 3$\sigma$ and Bonferroni gates at
macro AUC $= 0.70$, $z = +8.0$, $p = 1.1{\times}10^{-15}$ across four
training regions; a window-shift control and a foreshock-period mask
both confirm the signal lives exclusively in the final 5 days of the
pre-registered window. Per-target-magnitude stratification yields a clean
monotonic scaling: macro AUC $= 0.62$ for $M{\in}[4.5,5.0)$ vs $0.81$
for $M{\ge}5.0$. The result is a rigorous cross-regional confirmation of
the pervasive-foreshock phenomenon documented per-region by
\citet{trugman2019}, with a quantitative AUC and the first
M-scaling estimate under pre-registration discipline.
\end{abstract}

% =============================================================================
\section{Introduction}
\label{sec:intro}

The earthquake precursor literature has been historically constrained by
two structural problems: most claims are per-region (no cross-regional
generalization test) and post-hoc (no committed-in-advance protocol). The
Collaboratory for the Study of Earthquake Predictability
\citep{schorlemmer2007,mizrahi2024} solved pre-registration for
\emph{rate}-based forecasts; recent ionospheric-precursor work
\citep{tao2025iono} ported the discipline to a different physical domain.
No equivalent exists for \emph{feature}-based seismic precursor work.

In parallel, the foreshock literature has converged on the empirical
finding that elevated seismicity precedes most moderate mainshocks.
\citet{trugman2019} used a 1.81-million-event high-resolution
template-matched catalog to show that 72\% of $M{\ge}4$ mainshocks in
southern California have foreshock activity statistically elevated above
local background. The phenomenon is theoretically anchored by the ETAS
cascade-triggering framework \citep{helmstetter2003a, helmstetter2003b},
which derives expected foreshock-rate properties (acceleration toward
mainshock, magnitude-dependence, modified $b$-value) from a single
self-similar triggering process without invoking a distinct precursor
mechanism. Recent work continues to refine the methodology
\citep{lippiello2025, khan2025, moutote2021}.

We bridge these two threads. We pre-registered a CSEP-style six-feature
cross-regional evaluation on declustered $M{\ge}4.5$ targets across
six training regions (California, Cascadia, Japan, Chile, Turkey, Italy)
and two held-out test regions (Mexico, Alaska); the locked protocol
\citep{loftus2026prereg} commits feature definitions, region polygons,
station list, null models, magnitude-of-completeness procedure, evaluation
metric, and significance thresholds before any AUC was computed. One
amendment \citep{loftus2026pra2} re-ran the evaluation after v1 surfaced
two execution-time failure modes. The full chain of custody is publicly
available as a single repository with one commit per session.

This paper reports two contributions. First, a five-failure-mode
catalogue of how pre-registered cross-regional precursor evaluation
breaks in practice; each mode is documented with diagnostic evidence and
attempted mitigation. Second, a rigorous cross-regional confirmation of
the \citet{trugman2019} foreshock-pervasiveness finding using a TLS-style
template-correlation scan over the Benioff strain trajectory: macro AUC
$= 0.70$ at $z = +8.0$, with quantitative magnitude-scaling. We frame
this as confirmation rather than discovery: the foreshock phenomenon is
established, but the cross-regional AUC + M-scaling under explicit
pre-registration is novel.

% =============================================================================
\section{Pre-Registration Protocol}
\label{sec:prereg}

The locked v1 protocol \citep{loftus2026prereg} commits: ANSS ComCat
catalog 2000--2024 ingested at $M{\ge}1.5$; per-region $M_c$ via
Wiemer-Wyss \citep{wiemer2000} max-curvature $+$ Woessner-Wiemer
\citep{woessner2005} $+0.20$ correction $+$ explicit $b$-vs-$M_c$
plateau check; ZBZ 2013 nearest-neighbor declustering
\citep{zaliapin2013} with $b{=}1.0$, $d_f{=}1.6$,
$\log_{10}\eta_0{=}{-}5.0$; declustering applied to both the catalog and
the target set; six features (b-value, b-drift, Benioff total $+$
curvature, $n$ above $M_c$, plus three waveform-derived: spectral slope,
whitened-spectrogram entropy, HHT IMF1 instantaneous frequency); three
null types (A: aftershock-free random; B: minimal exclusion; C:
colored-noise synthetic for waveform features only); macro-AUC across
qualifying training regions with cross-region bootstrap CI95,
within-region permutation $z$, and Bonferroni-corrected $\alpha$. Pre-reg
v1 also commits an explicit ``will not'' list (no $M_c$ tuning, no
station/region swaps, no feature additions, no window changes).

Amendment v2 \citep{loftus2026pra2}, committed before re-evaluation,
makes three changes grounded in v1 execution evidence: ISC global
bulletin replaces ANSS ComCat (fixing failure mode~\#2 below); the
overlap-rejection forbidden zone shrinks from $[t'-30, t'+60]$ to
$[t', t'+60]$ (partial fix for failure mode~\#1 below); a per-region
minimum of $N{\ge}8$ kept precursor windows is required for inclusion in
the macro pool. The chain is binding: v1 SHA \texttt{a4f1c6f},
amendment SHA \texttt{05a4b0f}, all subsequent commits cite both.

% =============================================================================
\section{Methods}
\label{sec:methods}

\subsection{Catalog and station selection}
\label{sec:methods-catalog}

We use the ISC global bulletin via \texttt{obspy.clients.fdsn} for all
six training and two test regions over 2000-01-01 to 2024-12-31.
Catalog ingestion at $M{\ge}1.5$ as pre-registered would have been
intractably large at ISC; we use $M{\ge}2.5$ as a declared deviation, an
empirical compromise between bulletin-pull cost and $M_c$ identifiability.
Per-region $M_c$ is the chosen value where the $b$-vs-$M_c$ curve
plateaus to $|\Delta b| < 0.02$ between consecutive $M_c$; values are
recorded as artifacts and not retuned. Per-region waveform routing was
not pre-specified: we use NCEDC for California (BK network, Berkeley-
hosted), IRIS for Cascadia (UW.LON), Turkey (IU.ANTO), and Italy. Italy's
pre-registered station IV.AQU is gone post-2009; we use MN.AQU on IRIS,
the post-2009 MedNet successor at the same physical L'Aquila Observatory,
documented as 95\%-availability fallback per pre-reg \S 3.

\subsection{Per-window scalar features}
\label{sec:methods-features}

All five catalog-derived scalars and three waveform-derived scalars are
computed per 30-day pre-event window per pre-reg \S 4. The waveform
features are aggregated across six 10-minute snapshots evenly spaced
across the window using the region's primary broadband Z channel,
preprocessed (instrument-response-removal to ground velocity, bandpass
0.5--12~Hz clamped to $f_s/3$ to dodge any anti-alias artifacts).
Per-snapshot Welch PSD log-log slope (Brune 1970), whitened-spectrogram
Shannon entropy \citep{bensen2007}, and EMD IMF1 instantaneous frequency
\citep{huang1998} are aggregated to per-window scalars by mean (slope,
entropy) and median (IMF1 IF). Repeating-event rate, per-pre-reg
\S 4.1, is excluded from the headline panel as deferred infrastructure.

\subsection{Joint multi-feature classifier}
\label{sec:methods-clf}

In addition to per-feature scalar AUC, we train a joint classifier over
the six computable features (the $b$-pair excluded due to failure
mode~\#3 below) under leave-one-region-out (LORO) cross-validation. Each
region's features are z-scored within region using the union of
precursor and Null A windows in that region (per-region domain
adaptation); pooled samples from three training regions are used to fit a
logistic regression (\texttt{class\_weight="balanced"}) and a 200-tree
random forest, then scored on the held-out region.

\subsection{TLS-style template-correlation scan}
\label{sec:methods-tls}

Pre-reg \S 3.2 promised a TLS-style matched filter applied to feature
time series. We realize this via per-window 6-point trajectories at
5-day sub-window resolution. Templates are LORO precursor-mean
trajectories (averaged across the three training regions); each held-out
window is scored by Pearson correlation with the template; ROC-AUC is
computed precursor vs Null~A.

\subsection{Significance gates}
\label{sec:methods-gates}

Every macro statistic is evaluated against three concurrent gates per
pre-reg \S 6.3: cross-region bootstrap CI95-lower exceeds 0.5;
$|AUC - 0.5|$ exceeds 0.05; within-region permutation $z$-score exceeds
3 in magnitude. Bonferroni correction is applied per pre-reg \S 6.4 with
denominator equal to (\# features) $\times$ (\# nulls) $\times$
(\# qualifying regions); for the joint classifier and TLS analyses
the denominators are 2 and 8 respectively.

% =============================================================================
\section{Results}
\label{sec:results}

\subsection{Five-failure-mode catalogue}
\label{sec:results-failure-modes}

\paragraph{Failure mode \#1: overlap rule $\times$ event density.} The
pre-registered overlap rule was calibrated against California-scale
event density. Under v1's $[t'-30, t'+60]$ forbidden zone, regions with
moderate-to-high target density lose nearly all precursor windows: Japan
(4{,}563 declustered $M{\ge}4.5$ targets) loses all of them; Chile
(3{,}275) loses all; Cascadia (277) keeps only 4; Turkey (389) keeps 2.
Amendment 2 shrinks the zone to $[t', t'+60]$ and recovers Cascadia to
36 and Turkey to 30 kept precursor windows; Japan and Chile remain
inaccessible (1 each).

\paragraph{Failure mode \#2: ANSS ComCat $\times$ non-US $M_c$.} Under
v1's ANSS ComCat catalog, Japan $M_c = 4.55$ and Chile $M_c = 4.45$, both
at the $M{\ge}4.5$ target threshold. b-value features become structurally
uncomputable. Amendment 1 switches to ISC and recovers Japan/Chile $M_c$
to 3.5; California $M_c$ rises slightly (2.75$\to$3.50, an ISC US-coverage
gap relative to NCEDC) but California still qualifies.

\paragraph{Failure mode \#3: $b$-feature $\times$ $M_c$-window
incompatibility.} At PRA-2 $M_c$ values combined with 30-day windows, the
median $n$ of events $\ge M_c$ is 4--5 in California, Cascadia, and Italy,
and 76 in Turkey. The $b$-value MLE's $N{\ge}30$ floor is therefore
satisfied in only 3\% of windows for the first three regions and 73\% in
Turkey. The \citet{gulia2019} $b$-drop signal does not transfer to short
windows at moderate $M_c$. The $b$-vs-Null A macro AUC of 0.722 reported
elsewhere in the panel is a degenerate-sample artifact (Cascadia AUC =
1.00 computed on 1 precursor $\times$ 2 null finite values).

\paragraph{Failure mode \#4: test-region inaccessibility.} Both
pre-registered test regions, Mexico (843 declustered targets) and Alaska
(847), reduce to one kept precursor window each under the PRA-2 overlap
rule, far below the $N{\ge}8$ minimum. The cross-regional test-region
claim is therefore structurally inaccessible under any reasonable rule.
This is failure mode \#1 generalized to held-out regions. Test-region
strengthening of any precursor upper bound is impossible without further
amendments addressing subduction-zone seismicity density.

\paragraph{Failure mode \#5: precursor-window-includes-foreshock-period.}
The pre-reg defines the precursor window as $[t-30, t)$, including the
foreshock period. Any signal detected by trajectory-shape matching may
therefore reflect foreshock-period seismicity rather than 25-day-ahead
long-distance precursor. We address this with two controls
(\S\ref{sec:results-controls}); the result is that under the pre-reg's
literal definition, our positive finding is a real precursor signal,
but operationally it is foreshock detection.

\subsection{Per-window scalar and joint-classifier results}
\label{sec:results-scalar}

Across the eight-feature $\times$ two-null panel evaluated on the four
qualifying training regions (California, Cascadia, Turkey, Italy), zero
of sixteen tests pass 3$\sigma$ or Bonferroni at $\alpha = 0.05/128$.
Computable catalog features (Benioff total, Benioff curvature, $n$ above
$M_c$) lie within 0.05 of 0.5; waveform features (slope, entropy, IMF1
IF) lie 0.45--0.49 vs Null A with consistent direction across regions
but no individual feature reaching 3$\sigma$. The $b$ and $b$-drift
features are uncomputable (failure mode \#3) and their AUCs are
degenerate.

The joint LR + RF classifier under LORO yields macro AUC $= 0.526$ (LR)
and $0.480$ (RF), with permutation $z = +0.97$ and $-0.71$ respectively;
neither passes 3$\sigma$ or Bonferroni. The combination of features
carries no signal that the per-feature scalars miss.

\subsection{TLS template-correlation scan}
\label{sec:results-tls}

The TLS scan over the Benioff strain trajectory is the project's only
positive cross-regional finding. Macro AUC $= 0.704$, cross-region
bootstrap CI95 $= [0.637, 0.773]$, permutation $z = +8.01$, $p =
1.1 \times 10^{-15}$. Per-region (LORO held out): California 0.733,
Cascadia 0.809, Turkey 0.608, Italy 0.665; all four splits above 0.5,
direction-consistent. The result passes both the 3$\sigma$ pre-reg gate
and the Bonferroni-corrected $\alpha = 0.05/8 = 6.25 \times 10^{-3}$. The
$n$-above-$M_c$ trajectory yields macro AUC $= 0.542$ ($z = +1.62$),
sub-3$\sigma$. The waveform-feature trajectories were not TLS-scanned
(per-window aggregation only; deferred follow-up).

\subsection{Window-shift and foreshock-mask controls}
\label{sec:results-controls}

We applied two controls to interpret the TLS positive. First, the
precursor windows were shifted 30 days earlier ($[t-60, t-30)$, no
target overlap) while null windows were left untouched. Macro AUC
collapses from 0.704 to 0.535 ($z = +1.30$); the signal does not
generalize to a 30-day-shifted window. Second, the trajectory was
truncated to the first five sub-windows (days 0--25, foreshock period
masked). Macro AUC collapses to 0.513 ($z = +0.49$); per-region AUCs
drop to within 0.05 of 0.5 for all four regions.

The conclusion is unambiguous: the TLS positive is exclusively
foreshock-driven. The literal-pre-reg pass at AUC $= 0.70$ reflects the
final 5 days of the pre-registered window, where foreshock activity
dominates the M-weighted Benioff strain trajectory.

\subsection{Quantitative magnitude scaling}
\label{sec:results-magnitude}

Stratifying the 139 precursor windows by target magnitude yields a
clean monotonic scaling. For $M \in [4.5, 5.0)$ (n = 91), macro AUC
$= 0.624$. For $M \ge 5.0$ (n = 48), macro AUC $= 0.813$ — all four LORO
splits above 0.7 (California 0.81, Cascadia 0.85, Turkey 0.72, Italy
0.876). The signal scales monotonically with target $M$, consistent
with the ETAS cascade-triggering prediction \citep{helmstetter2003a}
that larger mainshocks correlate with stronger pre-event triggered
seismicity.

% =============================================================================
\section{Discussion}
\label{sec:discussion}

\subsection{What the failure-mode catalogue contributes}

Each of the five failure modes is a concrete, generalizable lesson for
future cross-regional pre-registered work in this area. Modes \#1 and
\#4 are structural mismatches between rate-forecast-derived overlap
conventions and the precursor-window framework; mode \#2 is an artifact
of relying on a US-centric catalog for international regions; mode \#3
is a sample-size problem at moderate $M_c$ with short windows; mode
\#5 is a definition issue with the precursor-window itself. None of
these is novel as an individual observation, but the catalogued
combination is a reusable artifact for the next group attempting the
same kind of work.

\subsection{Cross-regional confirmation of foreshock pervasiveness}

The TLS positive is a quantitative cross-regional confirmation of the
\citet{trugman2019} southern-California finding: foreshock activity is
elevated before most $M{\ge}4.5$ mainshocks, and the elevation is
detectable at AUC = 0.7 cross-regionally with monotonic M-scaling. The
particular metric (macro AUC of Benioff-strain trajectory template
correlation) and the cross-regional pre-registration protocol are novel,
but the underlying physical phenomenon is established. The
\citet{helmstetter2003a, helmstetter2003b} ETAS framework explains the
monotonic M-scaling we measured: larger mainshocks trigger more
preceding cascades.

\subsection{What the foreshock-period concentration means}

The window-shift and mask controls together establish that the entire
TLS signal lives in the final 5 days of the 30-day pre-registered
window. This is operationally important: the result is foreshock
detection, not 25-day-ahead long-distance precursor. The pre-reg's
literal definition of ``precursor window'' as $[t-30, t)$ admits the
foreshock period; under the literal protocol, the pass is real, but
under any interpretation that distinguishes foreshocks from
long-distance precursors, the long-distance signal is null.

A reasonable extension of this work would re-run with a precursor
window that explicitly excludes the final 5--7 days, testing whether any
sub-detection-threshold long-distance signal exists. We did not run this
because it would be a post-hoc deviation from pre-reg v1; we flag it as
the natural follow-up.

\subsection{Score honest, not inflated}

A natural temptation after detecting the TLS positive is to claim a
``first cross-regional precursor finding.'' We resist this. The
foreshock-pervasiveness phenomenon is well-documented per-region
\citep{trugman2019, lippiello2025, khan2025, hirose2021}; our work
adds quantitative cross-regional AUC and M-scaling under pre-registration
discipline but does not discover new physics. The honest contribution is
methodological: the failure-mode catalogue, the pre-registration
template, and the cross-regional AUC + M-scaling estimate.

% =============================================================================
\section{Conclusion}
\label{sec:conclusion}

We pre-registered a CSEP-style cross-regional precursor evaluation and
ran it under the locked protocol with one documented amendment. Five
methodological failure modes surfaced in execution, all catalogued with
diagnostic evidence. The per-feature scalar and joint-classifier
analyses yield clean nulls. The pre-registered TLS-style template
scan over Benioff strain trajectory passes 3$\sigma$ and Bonferroni at
macro AUC $= 0.70$, $z = +8$ across four training regions; this signal
is exclusively foreshock-period and scales monotonically with target
magnitude. We frame the result as cross-regional confirmation of
\citet{trugman2019} pervasive-foreshock activity with quantitative AUC
and the first M-scaling estimate under pre-registration discipline,
not as a novel discovery. The five-failure-mode catalogue is the
methodological contribution to future cross-regional pre-registered
precursor work.

\section*{Acknowledgements}
The author thanks the ISC for maintaining the global bulletin, the
NCEDC for hosting California waveform data, and INGV/MedNet for the
post-2009 successor station MN.AQU. ObsPy \citep{beyreuther2010} and
\texttt{usgs-libcomcat} provided the data-access infrastructure.
Pre-registration drafting and post-detection control design used
Anthropic Claude as an interactive collaborator (Claude is acknowledged
but not a co-author per arXiv policy).

\section*{Data and code availability}
All code, the pre-registration document \citep{loftus2026prereg}, the
amendment \citep{loftus2026pra2}, per-region cached catalogs, per-window
feature CSVs, and reproducible experiment scripts are publicly available
at \citet{loftus2026repo}. Each session of work is a single git commit;
the chain of custody from pre-reg through results is verifiable
without access to any private artifact.

\bibliography{references}
\end{document}
